---
title: "Effects of water, sanitation, and hygiene interventions on detection of enteropathogens and host-specific faecal markers in the environment: an individual-participant data meta-analysis"
author: "Andrew Mertens, Benjamin F. Arnold, Jade Benjamin-Chung, Alexandria Boehm, Joe Brown, Drew Capone, Thomas Clasen, Erica Fuhrmeister, Jessica Grembi, David Holcomb, Jackie Knee, Laura H Kwong, Audrie Lin, Stephen P. Luby, Rassul Nala, Kara Nelson, Sammy Njenga, Clair Null, Amy J. Pickering, Mahbubur Rahman, Heather Reese, Lauren Steinbaum, Jill Stewart, Ruwan Thilakaratne, Oliver Cumming, John M. Colford Jr., Ayse Ercumen"
#output: redoc::redoc
output: 
  officedown::rdocx_document:
      reference_docx: "manuscript_format.docx"
header-includes:
  - \setlength{\parindent}{4em}
  - \setlength{\parskip}{0em}
indent: true
bibliography: WASH-IPD.bib
bib-humanities: FALSE
csl: nature.csl

---

```{r, include=F}
#Note: Check out this package for integrating any edits from Ayse:
#https://github.com/noamross/redoc
#CSL styles by journal
#https://github.com/citation-style-language/styles
knitr::opts_chunk$set(echo = FALSE, dpi=300,fig.width=7)
options(warn = -1) 
library(officedown)
#library(janitor)
library(knitr)
library(here)
# library(table1)
# library(rvest)
library(officer)
library(flextable)
library(bibtex)
library(citr)
library(redoc)
source(here("0-config.R"))
#citr:::insert_citation(bib_file ="WASH-IPD.bib", use_betterbiblatex =F)
# 
# tidy_bib_file(
#   rmd_file = "Aim1_manuscript.Rmd"
#   , messy_bibliography = "WASH-IPD.bib"
#   , file = "tidy_references.bib"
# )
op = function(x, d=2) sprintf(paste0("%1.",d,"f"), x) 
Print <- function(x){
  prettyNum(round(x,2), big.mark = ",", scientific = FALSE)
}
 library(pander)
panderOptions('digits', 2)
panderOptions('round', 2)
panderOptions('keep.trailing.zeros', TRUE)
panderOptions('graph.fontsize', 8)
panderOptions('table.alignment.default', function(df)
    ifelse(sapply(df, is.numeric), 'right', 'left'))
panderOptions('table.split.table', Inf)
panderOptions('big.mark', ",")
set_flextable_defaults(big.mark = ",", 
  font.size = 8, theme_fun = theme_vanilla,
  padding.bottom = 5, 
  padding.top = 5,
  padding.left = 5,
  padding.right = 5)
#load clean data
d <- readRDS(paste0(dropboxDir,"Data/cleaned_ipd_env_data.rds"))
#sample-positive specific dataset
pos <- d %>% filter(!grepl("Any",sample),!grepl("any",sample),!grepl("Any",target), !is.na(pos))
#aggregate samples
pos.agg <- d %>% filter(grepl("Any",sample),grepl("any",sample),grepl("Any",target), !is.na(pos))
#abundance dataset
abund <- d %>% filter(!grepl("Any",sample),!grepl("any",sample),!grepl("Any",target), !is.na(abund))
#number of samples by study
num_samples_by_study <- d %>% filter(!grepl("Any",sample),!grepl("any",sample),!grepl("Any",target), !is.na(pos)) %>% 
  group_by(study, sample, clusterid, dataid, round, sampleid) %>% slice(1) %>%
  group_by(study) %>%
  summarize(N=n()) %>% mutate(study=gsub(" et al.","",as.character(study))) %>% arrange(N)
#number of food prep samples 
FP <- d %>% filter(sample=="FP") %>% group_by(target)
#zoonotic prev by animals
d_animals <- d %>% filter(!is.na(animals), target %in% c("Any zoonotic","Any non-zoonotic")) %>%
  mutate(tr=animals) %>% droplevels()
dzoo <- d_animals %>% filter(target=="Any zoonotic")
animal_zoonotic_tab <- round(prop.table(table(dzoo$animals, dzoo$pos ),1)*100,1)
#Load figures and tables
load(here("figures/all_tables.Rdata"))
# load(here("figures/agg_tables.Rdata"))
load(here("figures/abundance_tables.Rdata"))
load(here("figures/all_figures.Rdata"))
load(here("figures/subgroup_figures.Rdata"))
load(here("figures/subgroup_figures_PD.Rdata"))
load(here("figures/abund_dist_figures.Rdata"))
#load(here("figures/density_figures.Rdata"))
load(here("figures/outcome_groups.Rdata"))
tab1 <- read.csv(here("data/tab1_template.csv")) %>% filter(!is.na(Number.of.samples))
colnames(tab1) <- c("Parent study",	"Study design", "Intervention", "Time between intervention and environmental sampling",	"Location",	"Author/ year", 	"Sample types",	"Targets",	"Analytic method",	"Number of samples")
adj_RR <- readRDS(file=here("results/adjusted_aim1_RR_pooled.Rds")) 
unadj_RR <- readRDS(file=here("results/unadjusted_aim1_RR_pooled.Rds")) 
zoo_RR <- adj_RR %>% filter(grepl("zoo",target))
adj_RR <- adj_RR %>% filter(!grepl("zoo",target))
pooled_adj_RR <- adj_RR %>% filter(study=="Pooled")
adj_RR <- adj_RR %>% filter(study!="Pooled")
adj_diff <- readRDS(file=here("results/adjusted_aim1_diff.Rds")) 
sig_diff <- adj_diff %>% ungroup() %>% filter(pval < 0.05, perc_ROQ >=50) %>% select(study, sample, target, coef, ci.lb, ci.ub, pval, perc_ROQ)
adj_emm <- readRDS(here("results/adjusted_aim1_emm.Rds"))
sig_emm <- adj_emm %>% filter(int.p<0.05, target %in% c("Any pathogen","Any MST"))
zoonotic_animals_RR <- readRDS(here("results/adjusted_zoonotic_animals.Rds"))
subgroup_RR <- readRDS(here("results/subgroup_aim1_RR_pooled.Rds"))
#get any pathogen P-values
trial_p <- subgroup_RR$trial_comp %>% filter(target=="Any pathogen", sample=="any sample type") 
urban_p <- subgroup_RR$urban_comp  %>% filter(target=="Any pathogen", sample=="any sample type") 
trial_p <- round(trial_p$pval,2)
urban_p <- round(urban_p$pval,2)
  
#Study-specific estimates
protective <- adj_RR %>% ungroup() %>% filter(!is.na(coef),RR<=1) %>% select(sample, target, RR, ci.lb, ci.ub, study)
n_protective <- paste0(round(nrow(protective)/nrow(adj_RR %>% filter(!is.na(coef))) * 100, 1),"% (",nrow(protective),"/",nrow(adj_RR %>% filter(!is.na(coef))),")")
sig <- adj_RR %>% ungroup() %>% filter(ci.lb<1 & ci.ub<1 | ci.lb>1 & ci.ub>1) %>% select(sample, target, RR, ci.lb, ci.ub, study)
n_per_sig <- paste0(round(nrow(sig)/nrow(adj_RR %>% filter(!is.na(coef))) * 100, 1),"% (",nrow(sig),"/",nrow(adj_RR %>% filter(!is.na(coef))),")")
sig_protective <- protective %>% ungroup() %>% filter(ci.lb<1 & ci.ub<1 | ci.lb>1 & ci.ub>1) %>% select(sample, target, RR, ci.lb, ci.ub, study)
n_per_sig_protective <- paste0(round(nrow(sig_protective)/nrow(adj_RR %>% filter(!is.na(coef))) * 100, 1),"% (",nrow(sig_protective),"/",nrow(adj_RR %>% filter(!is.na(coef))),")")
perc_no_pos = round(prop.table(table((adj_RR$minN==0)))*100,1)[2]
num_no_pos = paste0(as.character(sum(adj_RR$minN==0, na.rm=T)),"/",as.character(nrow(adj_RR)))
num_sparse_pos = paste0(as.character(sum(is.na(adj_RR$coef) & adj_RR$minN>0, na.rm=T)),"/",as.character(nrow(adj_RR)))
adj_sparse <- adj_RR %>% filter(minN!=0, is.na(coef), !grepl("zoo",target)) %>% mutate(inv_n = N-n)
too_neg=round(mean(adj_sparse$n <= 10 | ((adj_sparse$a < 2 | adj_sparse$c < 2) & adj_sparse$b > 1 & adj_sparse$d > 1)) * 100, 1)
too_pos=round(mean(adj_sparse$inv_n <= 10 | ((adj_sparse$b< 2 | adj_sparse$d < 2) & adj_sparse$a > 1 & adj_sparse$c > 1)) * 100, 1)
#N estimates
n_estimated <- paste0(sum(!is.na(adj_RR$coef)),"/",nrow(adj_RR))
#order prevalences and seperate by MST vs. pathogens
sample_target_prevs <- adj_RR %>% filter(!is.na(coef), !grepl("Any ",target), !grepl("any ",sample)) %>%
                      mutate(prev=round(((a+c)/N)*100,1)) %>% arrange(prev) %>%
                      select(study, sample, target, prev)
pathogen_sample_target_prevs <- sample_target_prevs %>% filter(!grepl("\\(",target))
MST_sample_target_prevs <- sample_target_prevs %>% filter(grepl("\\(",target))
adj_MST_RR  <- adj_RR %>% filter(grepl("MST", target))
round(prop.table(table(adj_MST_RR$RR<1))[2]*100,1)
#Pvalue adjustment
#min(p.adjust(adj_RR$pval, method="BH"), na.rm=T)
#difference between adjusted and unadjusted
d_unadj_comp<- left_join(adj_RR, unadj_RR, by=c("Y","sample","target","study"))
d_unadj_comp <- d_unadj_comp %>% mutate(log.diff= coef.x-coef.y, log.abs.diff=abs(coef.x-coef.y), diff= RR.x-RR.y, abs.diff=abs(RR.x-RR.y))
med_unadj_diff<-median(d_unadj_comp$log.abs.diff, na.rm=T)
#Clean results function
clean_est <- function(res, outcome="binomial"){
  if(outcome=="binomial"){
    
    est = format(round(res$RR, 2), nsmall = 2)
    ci.lb = format(round(res$ci.lb, 2), nsmall = 2) 
    ci.ub = format(round(res$ci.ub, 2), nsmall = 2)
    est.ci = paste0(est," (95% CI: ",ci.lb,", ",ci.ub,")")
  }else{
    est = format(round(res$coef, 2), nsmall = 2)
    ci.lb = format(round(res$ci.lb, 2), nsmall = 2) 
    ci.ub = format(round(res$ci.ub, 2), nsmall = 2)
    est.ci = paste0(est," (95% CI: ",ci.lb,", ",ci.ub,")")
  }
  return(est.ci)
}
clean_res  <- function(res, outcome="binomial"){
  if(outcome=="binomial"){
    res$est.ci = paste0("RR=",round(as.numeric(res$RR),1)," (95% CI: ",round(res$ci.lb,1),", ",round(res$ci.ub,1),")")
    res <- res %>% ungroup() %>%
      select(study, sample_type, target_f, est.ci, pval, a, b, c, d, N) 
    res$est.ci <- gsub("RR\\=1 \\(95\\% CI\\: NA, NA\\)","Not estimated",res$est.ci)
    
    res$pval <- as.character(round(res$pval,3))
    res$pval[is.na(res$pval)] <- ""
    res <- res %>% 
      rename(Study=study,
             Sample=sample_type,
             Target=target_f,
             Estimate=est.ci,
             `p-value`=pval, 
             `Positive,\nIntervention`=a,
             `Negative,\nIntervention`=b,
             `Positive,\nControl`=c,
             `Negative,\nControl`=d,
             `Total\nobservations`=N)
  }else{
    res$est.ci = paste0(round(as.numeric(res$coef),1),"Difference= (95% CI: ",round(res$ci.lb,1),", ",round(res$ci.ub,1),")")
    res <- res %>% ungroup() %>%
      select(sample_type, target_f, est.ci, pval, a, b, c, d, N) 
    res$est.ci <- gsub("Difference\\=0 \\(95\\% CI\\: NA, NA\\)","Not estimated",res$est.ci)
    res$pval <- as.character(round(res$pval,3))
    res$pval[is.na(res$pval)] <- ""
    res <- res %>% 
      rename(Study=study,
             Sample=sample_type,
             Target=target_f,
             Estimate=est.ci,
             `p-value`=pval, 
             `Positive,\nIntervention`=a,
             `Negative,\nIntervention`=b,
             `Positive,\nControl`=c,
             `Negative,\nControl`=d,
             `Total\nobservations`=N)
  }
  
  return(res)
}
clean_res_tab <- function(unadj_RR, adj_RR){
  unadj<-clean_res(unadj_RR) %>% select(Study,Sample,Target,Estimate, `p-value`) %>% 
    rename(`Unadjusted\nEstimate`=Estimate, `Unadjusted\np-value`=`p-value`) 
  adj<-clean_res(adj_RR)%>% rename(`Adjusted\nEstimate`=Estimate, `Adjusted\np-value`=`p-value`)
  
  res <- left_join(unadj, adj, by=c("Study", "Sample", "Target"))
  res <- res %>% arrange(Study,  Sample,  Target) %>% filter(Study!="Pooled")
  res$`Adjusted\nEstimate`[res$Study=="Odagiri 2016"] <- ""
  res$`Adjusted\np-value`[res$Study=="Odagiri 2016"] <- ""
  return(res)
}
FitFlextableToPage <- function(ft, pgwidth = 6){
  ft_out <- ft %>% autofit()
  ft_out <- width(ft_out, width = dim(ft_out)$widths*pgwidth /(flextable_dim(ft_out)$widths))
  return(ft_out)
}
add_footnote_with_rlang <- function(df, col, pattern, symbol){
  form_ <- sprintf("~ %s == '%s'", col, pattern)
  colname <- enquo(col)
  flextable::compose(
    x = df, j = col, i = as.formula(form_), 
    value = as_paragraph(
      as_chunk(!!colname), 
      as_chunk(symbol, props = fp_text(vertical.align = 'superscript')))
  ) 
}
add_footnote <- function(df, col, pattern, symbol){
  form_ <- sprintf("~ %s == '%s'", col, pattern)
  flextable::compose(
    x = df, j = col, i = as.formula(form_), 
    value = as_paragraph(pattern,
      as_chunk(symbol, props = fp_text(vertical.align = 'superscript')))
    ) 
}
#adj_RR %>% filter(target=="Any virus", sample!="any sample type") %>% select(sample, target, RR, ci.lb, ci.ub, study, a,b,c,d) %>% arrange(sample) %>% as.data.frame()
```

```{css, echo=FALSE}
span.comment-start{
    background: #e0f3db;
}
span.comment-end{
    background: #e0f3db;
}
span.comment-start::after{
    content: " (" attr(author) ", " attr(date) ") [";
}
span.comment-end::before{
    content: "]"
}
span.comment-text{
    background: #fdbb84;
}
```

```{r, echo=FALSE}
commentN <- 0
cmt <- function(txt, cm, author, date = "2021-01-01", comN = commentN){
  cmt_str <- paste0('<span class="comment-text">[',cm,']{.comment-start id="', comN, '" author="', author, '" date="', date, '"}', txt, '[]{.comment-end id="',commentN,'"}</span>')
  assign("commentN", commentN + 1, envir = .GlobalEnv)
  return(cmt_str)
}
```





## Figures


```{r, echo=F, warning=F, message=F, fig.width = 7.2, fig.height = 7, fig.align = 'left'}
p_adj_1
```

__Figure 1.__ Forest plots of WASH intervention  effects on the prevalence of any enteropathogen or type of enteropathogen (any bacteria, any virus, any protozoa and any STH) in different types of environmental samples. Pooled estimates are presented when there are four or more study-specific estimates for a specific sample type and target combination and are denoted with diamond-shaped points. Grey crossed points denote data that were too sparse to estimate a prevalence ratio (i.e., <10 positive observations). Samples of the same type from different locations (source vs. stored water, flies in kitchen vs. latrine, soil from courtyard vs. latrine) or different individuals (child vs. mother's hands) are plotted separately. Point estimates and confidence intervals are printed next to pooled estimates. All estimates are adjusted for potential confounders.


```{r, echo=F, warning=F, message=F, fig.width = 7.2, fig.height = 5}
p_adj_2
```

__Figure 2.__ Forest plots of WASH intervention effects on the prevalence of any MST marker or type of MST marker (human or animal MST markers) in different types of environmental samples. Pooled estimates are presented when there are four or more study-specific estimates for a specific sample type and target combination and are denoted with diamond-shaped points. Grey crossed points denote data that were too sparse to estimate a prevalence ratio (i.e., <10 positive observations). Samples of the same type from different locations (source vs. stored water, flies in kitchen vs. latrine, soil from courtyard vs. latrine) or different individuals (child vs. mother's hands) are plotted separately. Point estimates and confidence intervals are printed next to pooled estimates. All estimates are adjusted for potential confounders.


## Tables


### Table 1. Characteristics of included publications

```{r, echo=F, warning=F, message=F,  results = 'asis'}
FitFlextableToPage(flextable(tab1), pgwidth = 7)
```

### Table 2. Mean (SD) abundances of enteropathogen and MST targets by study arm. Means are log10-transformed gene copies for MST markers and mean egg counts for soil transmitted helminths (*Ascaris* and *Trichuris*). Intervention effects are shown as adjusted differences in log10-transformed gene copies and ratios of helminth egg counts between the intervention and control arms.

```{r, echo=F, results = 'asis'}
tab_adj_diff2 <- tab_adj_diff %>% 
  subset(., select=c("Study","Sample","Target", "N","% in ROQ","Control mean, median (SD)","Intervention mean, median (SD)", "Difference or ratio (95% CI)","P value","Wilcoxon P value"))
colnames(tab_adj_diff2)[8] <- "temp"
neg.binom_ests <- tab_adj_diff2$temp[grepl("\\*",tab_adj_diff2$temp)]
neg.binom_ests <- gsub("\\*","",neg.binom_ests)
tab_adj_diff2$temp <- gsub("\\*","",tab_adj_diff2$temp)
#colnames(tab_adj_diff2)[8] <- "Difference (95% CI)"
abund_tab <- FitFlextableToPage(flextable(tab_adj_diff2), pgwidth = 7)
for(i in 1:length(neg.binom_ests)){
  abund_tab <- add_footnote(abund_tab, col = 'temp', pattern = neg.binom_ests[i], symbol = 'a')
}
abund_tab <- set_header_labels(abund_tab, temp = "Intervention effect (95% CI)")
abund_tab
```
ROQ: Range of quantification; SD: Standard deviation; CI: Confidence interval; Wilcoxon P-value: Non-parametric Wilcoxon rank sum test P-value.

^a^ Marks ratio estimates from negative binomial models.


## Supplementary Figures

```{r, echo=F, warning=F, message=F, fig.width = 7.2, fig.height = 10}
p_adj_s1
```

__Figure S2.__ Forest plots of intervention effects on the prevalence of specific pathogens.


```{r, echo=F, warning=F, message=F, fig.width = 7.2, fig.height = 9}
p_adj_s2
```

__Figure S3.__ Forest plots of intervention effects on the prevalence of specific MST markers.





```{r, echo=F, warning=F, message=F, fig.width = 7.2, fig.height = 3}
p_abund_dist_capone           
p_abund_dist_holcomb           
p_abund_dist_fuhrmeister       
p_abund_dist_kwong            
p_abund_dist_steinbaum       
```

__Figure S4.__ Distributions of abundance outcomes. The X-axes are displayed on the log-10 scale. Black vertical lines mark the means, and dashed lines mark the medians. Values below the limit of detection were imputed with with half the limit of detection and values below the limit of quantification were imputed with the midpoint between the limits of detections and quantification, leading to some bimodal distributions.



```{r, echo=F, warning=F, message=F, fig.width = 7.2, fig.height = 6}
p_wet_1_PD
```


__Figure S5.__ 

Forest plots of any enteropathogen prevalence differences or any MST prevalence differences between intervention and control arms, stratified by whether the sample was collected during the wet versus dry season (defined by the 6 months of highest average rainfall).  Significant effect modification, as determined by the p-values on the regression model interaction term, is marked above points with asterisks (P < 0.05 = "\*", P < 0.01 = "\*\*", P < 0.001 = "\*\*\*"). Grey crossed points denote data that were too sparse to estimate a prevalence ratio (i.e., <10 positive observations).

  

```{r, echo=F, warning=F, message=F, fig.width = 7.2, fig.height = 6}
p_animals_1_PD
```

__Figure S6.__ Forest plots of any enteropathogen prevalence differences or any MST prevalence differences between intervention and control arms, stratified by whether any animals were present in the compound. Significant effect modification, as determined by the p-values on the regression model interaction term, is marked above points with asterisks (P < 0.05 = "\*", P < 0.01 = "\*\*", P < 0.001 = "\*\*\*"). Grey crossed points denote data that were too sparse to estimate a prevalence ratio (i.e., <10 positive observations).


```{r, echo=F, warning=F, message=F, fig.width = 7.2, fig.height = 6}
p_zoo_PD
```

__Figure S7.__ Forest plots of any enteropathogen prevalence differences or any MST prevalence differences between intervention and control arms, stratified by whether the pathogen is zoonotically transmitted. Grey crossed points denote data that were too sparse to estimate a prevalence ratio (i.e., <10 positive observations). Significant effect modification, as determined by the p-values on the regression model interaction term, is marked above points with asterisks (P < 0.05 = "\*", P < 0.01 = "\*\*", P < 0.001 = "\*\*\*"). Grey crossed points denote data that were too sparse to estimate a prevalence ratio (i.e., <10 positive observations).


```{r, echo=F, warning=F, message=F, fig.width = 7.2, fig.height = 7}
p_unadj_1
```

__Figure S8.__ Forest plots of unadjusted intervention effects on the prevalence of any enteropathogen or type of enteropathogen (any bacteria, any virus, any protozoa and any STH) in different types of environmental samples. Point estimates and confidence intervals are printed next to pooled estimates. Grey crossed points denote data that were too sparse to estimate a prevalence ratio (i.e., <10 positive observations).


```{r, echo=F, warning=F, message=F, fig.width = 7.2, fig.height = 7}
p_unadj_2
```

__Figure S9.__ Forest plots of unadjusted intervention effects on the prevalence of any MST marker or type of MST marker (human or animal MST markers) in different types of environmental samples. Point estimates and confidence intervals are printed next to pooled estimates. Grey crossed points denote data that were too sparse to estimate a prevalence ratio (i.e., <10 positive observations).

